---
title: "Fullmer_Exam_3"
author: "christian"
date: "11/26/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Exam 3
+ Which sample locations are functionally different from each other in terms of what C-substrates they can utilize?
+ Are Soil and Water samples significantly different overall (as in, overall diversity of usable carbon sources)? What about for individual carbon substrates?
+ If there are differences between samples, which C-substrates are driving those differences?
+ Does the dilution factor change any of these answers?
+ Do the control samples indicate any contamination?




#Reading, cleaning, preliminary manipulation

Read-in using the "|" as the deliminating factor.

```{r, echo=FALSE, include=FALSE}

library(tidyverse)

biologData=read.delim("./BioLogData_Exam3.csv",sep = "|")

```





```{r,echo=TRUE, include=TRUE}


biologTidy= biologData %>%pivot_longer(cols=starts_with("Hr_"), names_to = "Hr",names_prefix = "Hr_", values_to = "Absorbance" )


#Across repetitions & dilutions - was there a response at given sample location dilution
biologTidy<-mutate(biologTidy,Rep=as.factor(Rep), Dilution=as.factor(Dilution))




```
##Which sample locations are functionally different from each other in terms of what C-substrates they can utilize?
```{r,echo=TRUE, include=TRUE, results='asis'}


biologTidy<-biologTidy%>%mutate(Response=if_else(Absorbance==0, "N", "Y"))

WiderBio<-biologTidy%>%pivot_wider(names_from= c(Rep,Hr, Dilution),values_from = Absorbance, values_fill =list(Absorbance=0
))

WiderBio<-WiderBio%>%mutate(Absorbance_Sum = rowSums(.[,c(5:31)]))

noResponseWider<-WiderBio%>%filter(Absorbance_Sum==0)%>%select(Sample.ID,Substrate)
yesResponseWider<-WiderBio%>%filter(Absorbance_Sum>0)%>%select(Sample.ID,Substrate)


noResponse<-biologTidy%>%mutate(Sample.ID=as.character(biologTidy$Sample.ID))%>%group_by(Sample.ID)%>%filter(Absorbance==0)%>%select(Substrate, Sample.ID,Response)

yesResponse<-biologTidy%>%mutate(Sample.ID=as.character(biologTidy$Sample.ID))%>%group_by(Sample.ID)%>%filter(Absorbance>0)%>%select(Substrate, Sample.ID,Response)

cantUtilize<-setdiff(noResponse,yesResponse)

testUtilize<-select(cantUtilize,Sample.ID, Substrate)%>%setdiff(noResponseWider,cantUtilize)




```

